---
title: "HUGEN 2072 Project 1"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# Instructions

Below is a tutorial consisting of a mostly-completed QC pipeline and some prompts.

Project 1 has three components:

-   **Scripting/coding**: read through this file and complete the pipeline by contributing your code *where indicated*.

-   **Narrative**: knit the completed .Rmd to a code-integrated HTML report of your QC process and decisions.

    -   Your report must answer all the questions and perform all the tasks indicated in the instructions/prompts given for each step.

    -   You must submit a successfully-knitted HTML file and the `.Rmd` in order to earn a passing grade for this project.

    -   Set up your report to flow like an actual document you might show your supervisor to summarize the QC project, emphasizing what was done at each step rather than the code used to do it (e.g., how many samples/SNPs were filtered out at a given step?)

    -   **Submit your .Rmd and HTML to Canvas**

-   **Presentation**: you will make a 10 min (maximum) recording of yourself with Panopto in which you summarize this QC pipeline. Instructions will given on Canvas.

In each Part of this project you will implement one of the main steps of cleaning genotyping array data in a small dataset (note that we left a few steps out). You'll be shown how to use functions from the `{GWASTools}` package, and you'll supply a little of your own code to complete the pipeline that's been provided. At each step, you may be asked to create a filter for the data, to address errors/discrepancies, etc. Then for each step you'll make a short report (a few sentences and figures) explaining what you did.

Note that "sample" and "scan" are used interchangeably throughout this assignment, but that these are NOT synonymous with subject/participant/individual/etc. In this assignment, "SNP" and "probe" are used interchangeably, but in other contexts you should always remember the distinction.

**Since the pipeline below is incomplete, some of the R code chunks are set to `eval=FALSE` so that knitting to HTML will succeed. You will need to change them to `eval=TRUE` as you progress so that your knitted document shows the results of running each chunk.**

**It is expected that you will copy this file somewhere to your directory on the CRC cluster so that you can use the RStudio Server, where all the packages you need have been installed already.**

### Acknowlegement

This work was done is parallel with Kash, Caylin, Vanessa, and Cecileigh. However, I take ultimate responsibility for the content of this assignment.

## Part 0 - Setup and Introduction to Annotation Data Frames

First, load the packages we'll be using.

```{r load_packages, message=F, eval=TRUE}

# Load the required packages for the project (Install them if necessary)

library("GWASTools", quietly = TRUE)
library("GWASdata", quietly = TRUE)
library("SNPRelate", quietly = TRUE)

# Close any open GDS files before proceeding (think of this as starting off with
# a 'blank slate')

showfile.gds(closeall = T, verbose = T)

```

The array data (genotypes, intensities, and BAF/LRR information) for this project are contained in several GDS files; accompanying SNP and scan annotation are also provided as an `.RData` file.

The probe/SNP annotation is stored in an object called a scan annotation data frame, and the sample-level annotation is stored in a scan annotation data frame. Each of these objects is similar to an ordinary data frame. More precisely, a data frame is embedded within each. There is also accompanying metadata. To access the data and metadata, the `{GWASTools}` package provides several functions. Three basic functions you should familiarize yourself with are `getAnnotation()`/`pData()` (extracts a standard data frame containing the annotation), `getMetadata()`/`varMetadata()` (extracts a data frame of variable descriptions), and `getVariableNames()`/`varLabels()` (extracts a vector of just the variable names). You can also view a variable in one of these objects by using `$`, as you would for an ordinary data frame. You can also add new variables, but you need to be careful to update the metadata as well. Since we will be adding new variables in the process of cleaning our data, the following chunk of code shows you how to do this. Then you will practice looking at the annotation, and your task is to answer some questions about this data set.

First, load the annotation data (`annotation_data.RData`) that was provided with this project. **You need to supply the path to this file.**

```{r load_annotation, eval=TRUE}

# Load the probe (SNP) and sample annotation data frames, which have been saved
# in an .RData file

path_to_annotation_data <-
  "/ix1/hugen2072-2026s/p1/annotation_data.RData"
load(path_to_annotation_data)

# List the loaded objects

ls()

```

Here's a toy example showing you how to work with annotation data frames. **Read this carefully, since you'll need to do this throughout the assignment.**

```{r annotation_data_frame_example, eval=TRUE}

### How to add a new variable to an annotation data frame ###

#   Make a new copy of the SNP annotation (just for this example)

snp_annodf_temp <- snpAnnot

#   Notice that printing the object isn't that helpful

snp_annodf_temp

#   So extract the annotation itself (this is the 'normal' data frame embedded
#   in the SNP annotation data frame object). Now you can see the SNP-level
#   annotation

snp_data_temp <- getAnnotation(snp_annodf_temp)
head(snp_data_temp)

#   Add a new column to the annotation (an integer dummy variable)

snp_data_temp$dummy <- 1:nrow(snp_data_temp)

#   Now update the scan annotation object itself by replacing the what was in
#   the annotation "slot"

pData(snp_annodf_temp) <- snp_data_temp

#   You can see that the new column has been added

head(pData(snp_annodf_temp))

#   Extract the metadata so that it can be updated

meta_temp <- getMetadata(snp_annodf_temp)
meta_temp

#   Add a new row titled "dummy" with an appropriate labelDescription (note that
#   the metadata is a data frame with named rows but one column)

meta_temp["dummy", "labelDescription"] <- "This is a dummy variable"

#   Finally, replace the old metadata with the new metadata

varMetadata(snp_annodf_temp) <- meta_temp

#   Look at the data and metadata to make sure there the dummy variable has been
#   added

head(getAnnotation(snp_annodf_temp))
getMetadata(snp_annodf_temp)

#   Remove the temp files

rm(snp_annodf_temp, snp_data_temp, meta_temp)

```

Now we use `GdsGenotypeReader()` and `GdsIntensityReader()` to open the gds files that contain the array data. Then use `GenotypeData()` and `IntensityData()` to combine them with the SNP and sample annotation, since the functions we'll be using later require this information to be "packaged" together.

```{r load_array_data, eval=TRUE}

#   Get the genotypes from a gds file

genofile <-
  system.file("extdata", "illumina_geno.gds", package = "GWASdata")
gds_geno <- GdsGenotypeReader(genofile)

#   Combine the genotypes with the sample and SNP annotation

genoData <-
  GenotypeData(gds_geno, snpAnnot = snpAnnot, scanAnnot = scanAnnot)
genoData

#   Get the intensity data from a gds file

qxyfile <-
  system.file("extdata", "illumina_qxy.gds", package = "GWASdata")
gds_qxy <- GdsIntensityReader(qxyfile)

#   Combine with the sample and SNP annotation

qxyData <-
  IntensityData(gds_qxy, snpAnnot = snpAnnot, scanAnnot = scanAnnot)
qxyData

#   Get the BAF/LRR data from a gds file

blfile <-
  system.file("extdata", "illumina_bl.gds", package = "GWASdata")
gds <- GdsIntensityReader(blfile)

#   Combine the BAF/LRR data with the sample and SNP annotation

blData <-
  IntensityData(gds, snpAnnot = snpAnnot, scanAnnot = scanAnnot)
blData

```

## Part 0 - Report

Using the functions introduced above to help you, briefly describe the dataset in a few sentences. Answer the following questions. Whatever code you use to help answer these questions can go in the empty chunk that follows.

-   What is the difference between the scanID and subjectID variables? Why are they both provided?

The scanID is the ID for each unique scan wile the subjectID is unique per subject. Both are provided because each subject may have multiple scans/samples that may be intended or unintended duplicates depending on the design of the study.

-   How many families are represented, and how many samples are from each family?

13

-   How many samples are of each sex?

F scans: 35
M scans: 42

-   How many samples are from each population group?

CEU:	49			
YRI:	28	

-   How many samples are there? How many people did they come from?

Samples: 77
People: 43

-   How many SNPs are there?

3300

-   Which chromosomes are represented? How many SNPs are on each?

21: 1000			
22: 1000			
23: 1000			
24: 100			
25: 100			
26: 100	

```{r part0_report}
library(tidyverse)

# What is the difference between the scanID and subjectID variables? Why are they both provided?
getMetadata(scanAnnot)
scanAnnot %>%
  getAnnotation() %>%
  head()

# How many families are represented, and how many samples are from each family?
scanAnnot %>%
  getAnnotation() %>% 
  group_by(family) %>% 
  summarise(
    n = n()
  ) %>% 
  count

# How many samples are of each sex?
n_scans <- scanAnnot %>%
      getAnnotation() %>%
      nrow()
n_F_scans <- scanAnnot %>% 
  getAnnotation() %>% 
  filter(sex == 'F') %>% 
  nrow()
n_M_scans <- scanAnnot %>% 
  getAnnotation() %>% 
  filter(sex == 'M') %>% 
  nrow()
print(
  paste0(
    "Total scans: ",
    n_scans
  )
)
print(
  paste0(
    "F scans: ",
    n_F_scans
  )
)
print(
  paste0(
    "M scans: ",
    n_M_scans
  )
)

# How many samples are from each population group?
scanAnnot %>% 
  getAnnotation() %>% 
  group_by(race) %>% 
  summarize(
    n = n()
  )

# How many samples are there? How many people did they come from?
scanAnnot %>% 
  getAnnotation() %>% 
  group_by(subjectID) %>% 
  summarize(
    n = n()
  ) %>% 
  count()

# How many SNPs are there?
n_snp <- snpAnnot %>% 
  getAnnotation() %>% 
  nrow()
print(
  paste0(
    "There are ",
    n_snp,
    " SNPs."
  )
)

# Which chromosomes are represented? How many SNPs are on each?
print("For all calls:")
snpAnnot %>% 
  getAnnotation() %>% 
  select(chromosome, snpID) %>% 
  unique() %>% 
  group_by(chromosome) %>% 
  summarize(
    n = n()
  )

```

## Part 1 - Missing Call Rate for Samples and SNPs

The basic `{GWASTools}` functions for missingness are `missingGenotypeBySnpSex()` and `missingGenotypeByScanChrom()`. `missingGenotypeBySnpSex()` returns a list of three objects: `missing.counts`, `scans.per.sex`, and `missing.fraction`. Read the help file for this function so that you understand what information is contained in these and how the Y chromosome is treated by sex (where sex is as defined in the scan annotation). Similarly, `missingGenotypeByScanChrom()` returns `missing.counts`, `snps.per.chr`, and `missing.fraction`.

There are four basic missingness calculations to be done, in the following order.

1.  Calculate `missing.n1` = fraction of genotype calls missing over all samples (except that annotated ‘female’ individuals are excluded for Y chr SNPs)

2.  Calculate `missing.e1` = missing call rate per sample over all SNPs (excluding SNPs with all calls missing)

3.  Calculate `missing.n2` = missing call rate per SNP over all samples whose `missing.e1` is less than `0.05`

4.  Calculate `missing.e2` = missing call rate per sample over all SNPs with `missing.n2` values less than `0.05`

We'll want to store the missingness information in the annotation data frames as we proceed.

Here's how to calculate and store `missing.n1`.

```{r missing.n1, eval=TRUE}

### Using missingGenotypeBySnpSex to calculate missing.n1 ###

#   First look at some of the results

miss <- missingGenotypeBySnpSex(genoData)
names(miss)
head(miss$missing.counts)
miss$scans.per.sex
head(miss$missing.fraction)

#   We should make sure the snpIDs are in the same order in the annotation and
#   in the missingness report (miss$missing.fraction); they are

allequal(snpAnnot$snpID, as.numeric(names(miss$missing.fraction)))

#   Then go ahead and add missing.n1 as a new annotation data column, and update
#   the metadata

snpAnnot$missing.n1 <- miss$missing.fraction
varMetadata(snpAnnot)["missing.n1", "labelDescription"] <-
  paste(
    "fraction of genotype calls missing over all samples, except that females are excluded for Y chr SNPs"
  )

### Explore the missing call rate per probes a little more

#   Look at the distribution of missingness for all the probes

hist(
  snpAnnot$missing.n1,
  ylim = c(0, 100),
  xlab = "SNP missing call rate",
  main = "Missing Call Rate for All Probes"
)

#   How many SNPs are missing a genotype for every single sample? Apparently
#   there are 151 such SNPs

sum(snpAnnot$missing.n1 == 1)

#   Store a vector containing the snpIDs of all the SNPs with missingness of
#   100%

snpexcl <- snpAnnot$snpID[snpAnnot$missing.n1 == 1]

```

Now we calculate and store `missing.e1`. The steps are described in the comments below.

```{r missing.e1, eval=TRUE}

### Using missingGenotypeByScanChrom to calculate missing.e1

#   Apply missingGenotypeByScanChrom to genoData, using the snp.exclude option
#   to exclude these SNPs (why does it make sense to exclude those SNPs?)

miss <- missingGenotypeByScanChrom(genoData, snp.exclude = snpexcl)

#   Make sure the order of the scanIDs in the output from
#   missingGenotypeByScanChrom is the same as the order of the scanIDs in the
#   annotation data frame

allequal(names(miss$missing.fraction), scanAnnot$scanID)

#   Make a new column in scanAnnot called missing.e1 and store the per-scan
#   missingness from missingGenotypeByScanChrom there

scanAnnot$missing.e1 <- miss$missing.fraction

#   Update the scanAnnot metadata

varMetadata(scanAnnot)["missing.e1", "labelDescription"] <-
  "fraction of genotype calls missing over all snps with missing.n1<1 except that Y chr SNPs are excluded for females"

# We can check summary statistics and make a histogram, too

summary(scanAnnot$missing.e1)
hist(
  scanAnnot$missing.e1,
  xlab = "Fraction of missing calls over all probes",
  main = "Histogram of Sample Missing Call Rate for All Samples"
)

print("For only calls MCR < 1:")
snpAnnot %>% 
  getAnnotation() %>% 
  filter(missing.n1 < 1) %>% 
  select(chromosome, snpID) %>% 
  unique() %>% 
  group_by(chromosome) %>% 
  summarize(
    n = n()
  )

```

Now, supply your code to calculate and store `missing.n2`. The steps you should follow are described in the comments below.

```{r missing.n2, eval=TRUE}

### Now that missing.n1 and missing.e1 have been calculated, we can calculate missing.n2

#   Recall that missing.n2 = call rate per SNP over all samples whose missing.e1
#   is less than 0.05.

#   Since no sample had missing.e1 > 0.05 in this dataset, we don't actually
#   have to exclude any samples.

#   However, let's proceed as if there were some samples with missing.e1 > 0.05.

#   Make a vector of the scanIDs for the samples with high missing.e1.

snp_call_rate_gt_0_05 <- scanAnnot$scanID[scanAnnot$missing.e1 > 0.05]

#   Apply the missingGenotypeBySnpSex function again, using the scan.exclude
#   argument.

miss <- missingGenotypeBySnpSex(genoData, scan.exclude = snp_call_rate_gt_0_05)

#   Store missing.n2 as a new column in snpAnnot.

snpAnnot$missing.n2 <- miss$missing.fraction

#   Update the metadata for snpAnnot!

#   The labelDescription for missing.n2 should be "fraction of genotype calls
#   missing over all samples with missing.e1 < 0.05 except that females are
#   excluded for Y chr SNPs"

varMetadata(snpAnnot)["missing.n2", "labelDescription"] <-
  "fraction of genotype calls missing over all samples with missing.e1 < 0.05 except that females are excluded for Y chr SNPs"

summary(snpAnnot$missing.n2)
hist(
  snpAnnot$missing.n2,
  xlab = "Fraction of missing calls over all samples w/MCR < 0.05",
  main = "Histogram of SNPs Missing Call Rate for All Samples w/MCR < 0.05"
)

```

Write code to calculate and store `missing.e2`. The steps you should follow are described in the comments below.

```{r missing.e2, eval=TRUE}

### Last, we calculate missing.e2

#   Recall missing.e2 = missing call rate per sample over all SNPs with
#   missing.n2 values less than 0.05

#   Store a vector with the snpIDs of all probes with missing.n2 >= 0.05

sample_call_rate_gte_0_05 <- snpAnnot$snpID[snpAnnot$missing.n2 >= 0.05]

#   Apply missingGenotypeByScanChrom to genoData again, using the snp.exclude
#   argument to exclude the SNPs with missing.n2 >= 0.05 (why do we do this?)

miss <- missingGenotypeByScanChrom(genoData, snp.exclude = sample_call_rate_gte_0_05)

#   Make a new column in scanAnnot called missing.e2 to store the missingness
#   per-scan you just calculated

scanAnnot$missing.e2 <- miss$missing.fraction

#   Update the scanAnnot metadata

#   The labelDescription should say "fraction of genotype calls missing over all
#   snps with missing.n2 < 0.05 except that Y chr SNPs are excluded for females"

varMetadata(scanAnnot)["missing.e2", "labelDescription"] <-
  "fraction of genotype calls missing over all snps with missing.n2 < 0.05 except that Y chr SNPs are excluded for females"

hist(
  scanAnnot$missing.e2,
  xlab = "Fraction of missing calls over all snps w/MCR < 0.05",
  main = "Histogram of Sample Missing Call Rate for All SNPs w/MCR < 0.05"
)

```

## Part 1 - Report

Summarize the filtering you've accomplished at this step. Include the following in your answer the following questions.

-   (First, make sure you've filled in the calculations of `missing.n2` and `missing.e2` in the chunks above.)

-   How many SNPs and samples were genotyped?

-   Briefly describe the purpose of doing this step in two rounds (why do we calculate both missingnesses twice?)

-   What were the median, mean, and maximum per-sample missingness (use `missing.e2`)? Make a histogram of `missing.e2`.

-   How many samples had no missing calls at all (use `missing.e2`)?

-   Are the samples generally "good"? How many should be excluded from downstream analyses?

-   How many SNPs had a 100% missing call rate (use `missing.n1`)?

-   What are the median, mean, and maximum per-SNP missingnesses (use `missing.n2`)? How many SNPs are missing no calls at all?

-   Make a histogram of `missing.n2`.

-   How many SNPs are filtered out and how many remain if you retain only SNPs with `missing.n2` \< 0.05?

-   Make a histogram showing `missing.n2` for the remaining SNPs.

```{r part1_report}

# How many SNPs and samples were genotyped?

print("Number of SNPs attempted to be genotyped:")
nrow(snpAnnot)
print("Number of SNPs genotyped:")
snpAnnot %>% 
  getAnnotation() %>% 
  filter(missing.n1 < 1) %>% 
  nrow()
print("Number of Samples attempted to be genotyped:")
nrow(scanAnnot)
print("Number of Samples genotyped:")
scanAnnot %>% 
  getAnnotation() %>% 
  filter(missing.e1 < 1) %>%
  nrow()

# Briefly describe the purpose of doing this step in two rounds (why do we calculate both missingnesses twice?)

print("A: We undergo two rounds of missingness alternating between sample-wise and snp-wise filtering.")
print("The first round is to filter based on overall probe-wise (per probe for all samples) and sample-wise (per sample across all probes) missingness.")
print("The second round repeats the task with a limit of missingess less than 5%.")
print("Probe-wise for round 2 ensures we have only probes from high quality samples.")
print("Sample-wise for round 2 ensures we have only samples from high quality probes.")

# What were the median, mean, and maximum per-sample missingness (use `missing.e2`)?

print("See table and histogram for per-sample missingness overview...")
scanAnnot %>% 
  getAnnotation() %>% 
  summarise(
    median = median(missing.e2),
    mean = mean(missing.e2),
    max = max(missing.e2)
  )

# Make a histogram of `missing.e2`.

hist(
  scanAnnot$missing.e2,
  xlab = "Missing Call Rate",
  main = "Frequency of Sample Missingness in High Quality SNPs"
)

# How many samples had no missing calls at all (use `missing.e2`)?

print("Samples with no missing calls:")
scanAnnot %>% 
  getAnnotation() %>% 
  filter(missing.e2 == 0) %>% 
  nrow()

# Are the samples generally "good"? How many should be excluded from downstream analyses?

print("Overall samples to be excluded:")
scanAnnot %>% 
  getAnnotation() %>% 
  filter(missing.e2 > 0.05) %>% 
  nrow()
print("Overall, it was found that the samples were high quality and none were excluded from downstream analysis.")

# How many SNPs had a 100% missing call rate (use `missing.n1`)?

print("Number of SNPs with 100% MCR:")
n_100_perc_missing_snps <- snpAnnot %>% 
  getAnnotation() %>% 
  filter(missing.n1 == 1) %>% 
  nrow()
n_100_perc_missing_snps

# 151 ... now I have to add a pokemon easter egg for the original 151
# ...placed at the bottom of this report

# What are the median, mean, and maximum per-SNP missingnesses (use `missing.n2`)? How many SNPs are missing no calls at all?

print("See table and histogram for per-SNP missingness overview...")
snpAnnot %>% 
  getAnnotation() %>% 
  summarise(
    median = median(missing.n2),
    mean = mean(missing.n2),
    max = max(missing.n2)
  )

# Make a histogram of `missing.n2`.

hist(
  snpAnnot$missing.n2,
  xlab = "Missing Call Rate",
  main = "Frequency of SNP Missingness in High Quality Samples"
)

# How many SNPs are filtered out and how many remain if you retain only SNPs with `missing.n2` \< 0.05?

print("Remaining SNPs if only retaining SNPs with MCR < 0.05")
retained_snps <- snpAnnot %>% 
  getAnnotation() %>% 
  filter(missing.n2 < 0.05)
retained_snps %>% nrow()

# Make a histogram showing `missing.n2` for the remaining SNPs.
retained_snps$missing.n2 %>% 
  hist(
    xlab = "Missing Call Rate",
    main = "Frequency of SNP Missingness in High Quality Samples w/MCR < 0.05"
  )

```

```{r}

retained_scans <- scanAnnot %>%
  getAnnotation() %>% 
  filter(missing.e2 < 0.05)

# How many families are represented, and how many samples are from each family?
retained_scans %>% 
  group_by(family) %>% 
  summarise(
    n = n()
  ) %>% 
  count

# How many samples are of each sex?
n_scans <- retained_scans %>%
      nrow()
n_F_scans <- retained_scans %>% 
  filter(sex == 'F') %>% 
  nrow()
n_M_scans <- retained_scans %>% 
  filter(sex == 'M') %>% 
  nrow()
print(
  paste0(
    "Total scans: ",
    n_scans
  )
)
print(
  paste0(
    "F scans: ",
    n_F_scans
  )
)
print(
  paste0(
    "M scans: ",
    n_M_scans
  )
)

# How many samples are from each population group?
retained_scans %>% 
  group_by(race) %>% 
  summarize(
    n = n()
  )

# How many samples are there? How many people did they come from?
retained_scans %>% 
  group_by(subjectID) %>% 
  summarize(
    n = n()
  ) %>% 
  count()

retained_snps %>% nrow()

retained_snps %>% 
  select(chromosome, snpID) %>% 
  unique() %>% 
  group_by(chromosome) %>% 
  summarize(
    n = n()
  )

```

## Part 2 - Inferred vs. Reported Sex and Relatedness

To investigate discrepancies in inferred vs. reported sex (and, possibly, sex chromosome aneuploidies), we will produce four plots that summarize each scan's heterozygosity and intensity for the sex chromosomes:

1.  Mean Y chromosome vs. mean X chromosome intensity

2.  Mean X chromosome heterozygosity vs. mean X chromosome intensity

3.  Mean Y chromosome intensity vs. mean X chromosome heterozygosity

4.  Mean X chromosome heterozygosity vs. autosomal heterozygosity

Recall that the intensity data was read in earlier as `qxyData`. Before making our graphs, we'll need to calculate the variables we want to plot. The basic function we'll use to do this is `meanIntensityByScanChrom()`, which will return a list of matrices. Read the help file of this function so that you understand the output. Heterozygosity is calculated with `hetByScanChrom()`.

```{r sex_check, eval=TRUE}

#   Look at the structure of the IntensityData object

qxyData

#   For each probe, there's a numerical ID, chromosome, position, rsID

#   The intensities (stored as variables called X and Y) are each essentially
#   3300×77 matrices (an X and Y intensity for each scan for every single probe)

#   (Don't confuse the *intensity variable names* _X_ and _Y_ in the IntensityData
#   object with the X and Y *chromosomes*!)

#   Apply the meanIntensityByScanChrom function to get the intensity averages.
#   (We're not going to use all of the output from this function)

inten.by.chrom <- meanIntensityByScanChrom(qxyData)

#   Extract the matrix containing the mean intensity for each chromosome for
#   each scan

names(inten.by.chrom)
mninten <- inten.by.chrom$mean.intensity

#   The matrix is 77x6 since there are 77 samples and 6 chromosomes represented
#   in the dataset

dim(mninten)
head(mninten)

#   We also need to calculate the X chromosome and autosome heterozygosities for
#   each sample before we can plot them

#   Apply hetByScanChrom to genoData (we need actual genotypes to calculate the
#   heterozygosity)

het.results <- hetByScanChrom(genoData)

#   The function calculates eachs scan's heterozygosity for each chromosome, and
#   also for the autosomes as a whole

#   We want to add X and autosome heterozygosity as new columns in scanAnnot

#   Make sure the scanIDs are in the right order first! (they are)

allequal(scanAnnot$scanID, rownames(het.results))

#   Extract the autosomal and X heterozygosity and store them in scanAnnot

scanAnnot$het.A <- het.results[, "A"]
scanAnnot$het.X <- het.results[, "X"]

#   Add appropriate metadata

varMetadata(scanAnnot)["het.A", "labelDescription"] <-
  "fraction of heterozygotes for autosomal SNPs"
varMetadata(scanAnnot)["het.X", "labelDescription"] <-
  "fraction of heterozygotes for X chromosome SNPs"

#   Before making the plots, there are a few things to do

#   We want to color the plotted points by annotated sex

#   But first, make sure that the order of the scanIDs (and hence the annotated
#   sexes) matches the order of the scanIDs in the data we'll plot

allequal(scanAnnot$scanID, rownames(mninten))

#   Now we assign each sex a color (male=blue, female=red)

#   Store the colors in a vector

xcol <- rep(NA, nrow(scanAnnot))
xcol[scanAnnot$sex == "M"] <- "blue"
xcol[scanAnnot$sex == "F"] <- "red"

#   We also want to count the number of SNPs getting included on the X and Y
#   chromosomes (which are coded as 23 and 25 in snpAnnot)

nx <- sum(snpAnnot$chromosome == 23)
ny <- sum(snpAnnot$chromosome == 25)

# Now make the plots

# All intensities

x1 <- mninten[, "X"]
y1 <- mninten[, "Y"]
main1 <- "Mean X vs \nMean Y Chromosome Intensity"

# Het on X vs X intensity

x2 <- mninten[, "X"]
y2 <- scanAnnot$het.X
main2 <-
  "Mean X Chromosome Intensity vs Mean X Chromosome Heterozygosity"

# Het on X vs Y intensity

y3 <- mninten[, "Y"]
x3 <- scanAnnot$het.X
main3 <-
  "MeanX Chromosome Heterozygosity vs Mean Y Chromosome Intensity"

# X vs A het

x4 <- scanAnnot$het.A[scanAnnot$sex == "F"]
y4 <- scanAnnot$het.X[scanAnnot$sex == "F"]
main4 <-
  "Mean Autosomal Heterozygosity vs Mean X Chromosome Heterozygosity"

# Make labels for axes/legends

cols <- c("blue", "red")
mf <- c("male", "female")
xintenlab <- paste("X intensity (n=", nx, ")", sep = "")
yintenlab <- paste("Y intensity (n=", ny, ")", sep = "")

# Make the 4 plots and add a legend

par(mfrow = c(2, 2))
plot(
  x1,
  y1,
  xlab = xintenlab,
  ylab = yintenlab,
  main = main1,
  col = xcol,
  cex.main = 0.8
)
legend("bottomleft",
       mf,
       col = cols,
       pch = c(1, 1),
       cex = 0.5)
plot(
  x2,
  y2,
  col = xcol,
  xlab = xintenlab,
  ylab = "X heterozygosity",
  main = main2,
  cex.main = 0.8
)
plot(
  x3,
  y3,
  col = xcol,
  ylab = yintenlab,
  xlab = "X heterozygosity",
  main = main3,
  cex.main = 0.8
)
plot(
  x4,
  y4,
  col = "red",
  xlab = "Autosomal heterozygosity",
  ylab = "X heterozygosity",
  main = main4,
  cex.main = 0.8
)
par(mfrow = c(1, 1))

```

## Part 2 - Report (Sex Check)

-   Describe any anomalies you see in the 4 plots above.

-   Figure out the scanIDs of any scans that look like outliers or possible instances of mislabeled sex. Include these IDs in your report and describe why they appear to be mislabeled.

-   For the purposes of this assignment we will ASSUME that some scans have just had their sexes mislabeled. (In "real life", we would not just assume this, and we might instead flag these scans for removal from some downstream steps/analyses.)

-   Manually change the sex annotation for these scans in scanAnnot to "remove" the discrepancy. (Manually changing your data is risky and not generally recommended. Be careful!). In your report, state exactly what changes you made. Add a new column to scanAnnot indicating this, too (add appropriate metadata). Does anything in the pedigree structure make you more confident that this "fix" is correct? (Are these samples labeled as mothers or fathers?)

-   Remake the 4 plots above, using the "corrected" data (nothing should appear mislabeled now for this dataset).

-   How might you follow up on this "in real life"?

```{r part2_report_sex_check}

# Describe any anomalies you see in the 4 plots above.

print("There appear to be 2 or 3 reported females that are inferring XY.")
print("These can be seen in the X vs Y intensity and heteroygosity plots and the X intensity vs X heterozygosity plots.")
print("There also appear to be some outliers with low X heterogeneity vs autosomal heteroygosity.")
print("These are likely the reported females that we can infer are male.")

# Figure out the scanIDs of any scans that look like outliers or possible instances of mislabeled sex. Include these IDs in your report and describe why they appear to be mislabeled.

# get the outliers from the reported females with a low X heterozygosity (arbitrarily chose 0.1 give the graphs)
diff_inferred_sex <- scanAnnot %>% 
  getAnnotation() %>% 
  filter(sex == "F" & het.X < 0.1) %>%
  select(scanID) %>%
  unlist()

print('Scans that appear to have mislabeled sex (in "incorrect:" cluster):')
diff_inferred_sex

# For the purposes of this assignment we will ASSUME that some scans have just had their sexes mislabeled. (In "real life", we would not just assume this, and we might instead flag these scans for removal from some downstream steps/analyses.)

# Manually change the sex annotation for these scans in scanAnnot to "remove" the discrepancy. (Manually changing your data is risky and not generally recommended. Be careful!). In your report, state exactly what changes you made. Add a new column to scanAnnot indicating this, too (add appropriate metadata). Does anything in the pedigree structure make you more confident that this "fix" is correct? (Are these samples labeled as mothers or fathers?)

# save the reported sex values
scanAnnot$reported_sex <- scanAnnot$sex
# update the descriptions of sex and reported_sex
varMetadata(scanAnnot)["sex", "labelDescription"] <-
  "sex coded as M=male and F=female as corrected from inference from the data"
varMetadata(scanAnnot)["reported_sex", "labelDescription"] <-
  "sex coded as M=male and F=female as reported in the raw data"

# manually update the sex given the scanIDs identified
updated_sex <- scanAnnot$sex
print("Before reported/inferred sex correction:")
table(updated_sex)
names(updated_sex) <- scanAnnot$scanID

for(scan in diff_inferred_sex) {
  idx <- toString(scan)
  updated_sex[idx] <- if_else(updated_sex[idx] == "F", "M", "F")
}

print("After reported/inferred sex correction:")
table(updated_sex)

scanAnnot$sex <- updated_sex

# Remake the 4 plots above, using the "corrected" data (nothing should appear mislabeled now for this dataset).

# can just copy & paste the graphing code above because we updated the "sex" column itself...
# --- start copied code ---

#   Apply hetByScanChrom to genoData (we need actual genotypes to calculate the
#   heterozygosity)

het.results <- hetByScanChrom(genoData)

#   The function calculates eachs scan's heterozygosity for each chromosome, and
#   also for the autosomes as a whole

#   We want to add X and autosome heterozygosity as new columns in scanAnnot

#   Make sure the scanIDs are in the right order first! (they are)

allequal(scanAnnot$scanID, rownames(het.results))

#   Extract the autosomal and X heterozygosity and store them in scanAnnot

scanAnnot$het.A <- het.results[, "A"]
scanAnnot$het.X <- het.results[, "X"]

#   Add appropriate metadata

varMetadata(scanAnnot)["het.A", "labelDescription"] <-
  "fraction of heterozygotes for autosomal SNPs"
varMetadata(scanAnnot)["het.X", "labelDescription"] <-
  "fraction of heterozygotes for X chromosome SNPs"

#   Before making the plots, there are a few things to do

#   We want to color the plotted points by annotated sex

#   But first, make sure that the order of the scanIDs (and hence the annotated
#   sexes) matches the order of the scanIDs in the data we'll plot

allequal(scanAnnot$scanID, rownames(mninten))

#   Now we assign each sex a color (male=blue, female=red)

#   Store the colors in a vector

xcol <- rep(NA, nrow(scanAnnot))
xcol[scanAnnot$sex == "M"] <- "blue"
xcol[scanAnnot$sex == "F"] <- "red"

#   We also want to count the number of SNPs getting included on the X and Y
#   chromosomes (which are coded as 23 and 25 in snpAnnot)

nx <- sum(snpAnnot$chromosome == 23)
ny <- sum(snpAnnot$chromosome == 25)

# Now make the plots

# All intensities

x1 <- mninten[, "X"]
y1 <- mninten[, "Y"]
main1 <- "Mean X vs \nMean Y Chromosome Intensity"

# Het on X vs X intensity

x2 <- mninten[, "X"]
y2 <- scanAnnot$het.X
main2 <-
  "Mean X Chromosome Intensity vs Mean X Chromosome Heterozygosity"

# Het on X vs Y intensity

y3 <- mninten[, "Y"]
x3 <- scanAnnot$het.X
main3 <-
  "MeanX Chromosome Heterozygosity vs Mean Y Chromosome Intensity"

# X vs A het

x4 <- scanAnnot$het.A[scanAnnot$sex == "F"]
y4 <- scanAnnot$het.X[scanAnnot$sex == "F"]
main4 <-
  "Mean Autosomal Heterozygosity vs Mean X Chromosome Heterozygosity"

# Make labels for axes/legends

cols <- c("blue", "red")
mf <- c("male", "female")
xintenlab <- paste("X intensity (n=", nx, ")", sep = "")
yintenlab <- paste("Y intensity (n=", ny, ")", sep = "")

# Make the 4 plots and add a legend

par(mfrow = c(2, 2))
plot(
  x1,
  y1,
  xlab = xintenlab,
  ylab = yintenlab,
  main = main1,
  col = xcol,
  cex.main = 0.8
)
legend("bottomleft",
       mf,
       col = cols,
       pch = c(1, 1),
       cex = 0.5)
plot(
  x2,
  y2,
  col = xcol,
  xlab = xintenlab,
  ylab = "X heterozygosity",
  main = main2,
  cex.main = 0.8
)
plot(
  x3,
  y3,
  col = xcol,
  ylab = yintenlab,
  xlab = "X heterozygosity",
  main = main3,
  cex.main = 0.8
)
plot(
  x4,
  y4,
  col = "red",
  xlab = "Autosomal heterozygosity",
  ylab = "X heterozygosity",
  main = main4,
  cex.main = 0.8
)
par(mfrow = c(1, 1))

# --- end copied code ---

# How might you follow up on this "in real life"?

print("I would check with any RAs to see if any surveys, general collection of the data, or the individual themselves may have specified a different gender identity than biological sex or if they maybe mislabeled the sex.")
print('If unable to confirm these inferences, I would consult the PI on if they want to keep these "corrected" data or drop these samples.')

scanAnnot %>% 
  getAnnotation() %>% 
  filter(scanID %in% diff_inferred_sex)

```

## Part 3 - Allelic Imbalance; BAF and LRR Plots

Here we want to look at B allele frequency (BAF) and log R ratio (LRR) plots using `chromIntensityPlot()`. Ordinarily, since there are 22 to 23 chromosomes to check for every sample, you would use a script that automatically detects aberrations in BAF and LRR. For this project, we'll pretend that our pipeline has already flagged one sample for a possible chromosomal anomaly.

```{r baflrr, eval=TRUE}

# Recall that earlier we read the intensity data into an IntensityData object
# called blData.

# The snp/scan annotation were attached as well.

# Suppose that chromosome 22 was flagged for scanID 286.

# Making the BAF/LRR plot is easy

#dev.off()
chromIntensityPlot(blData, scan.ids = 286, chrom.ids = 22)
par(mfrow = c(1, 1))

```

## Part 3 - Report

-   Describe the purpose of this step.

-   Show an example of a "normal" pair of BAF/LRR plots (choose any chromosome/scanID you want).

-   How would you detect trisomy 21 (i.e., describe briefly in words what the BAF/LRR plot would look like).

-   Suppose an individual had a balanced reciprocal translocation involving chromosome 2 and 4. What would you expect that individual's BAF/LRR plots would look like for those chromosomes?

-   Suppose chromosome 22 in sample 286 was the only chromosome flagged for a possible anomaly in this dataset. Looking at the plot, do you notice anything strange? Can you suggest what kind of anomaly might be showing up? (There isn't any single answer that's certainly right! The purpose of this question is to check that you understand what the BAF and LRR plots "mean").

-   Finally, suppose your pipeline has given you a list of the exact chromosomal segments showing possibly aneuploidies/anomalies in each sample. What would you recommend doing with the genotypes in those segments? (You're not being asked to implement this here - just say what you recommend doing.)

```{r part3_report}

# Describe the purpose of this step.

print("The purpose of this step is to check for a number of potential anomalies including:")
print("- segments of chromosomal deletions, duplicatons, and runs of homozygosity")
print("- potential aneuploidies")
print("- admixture, whole genome mosaicism, or chimerism")
print("Ultimitely, we are looking to genotypes to exclude or whole samples whose copy number is unexpected.")

# Show an example of a "normal" pair of BAF/LRR plots (choose any chromosome/scanID you want).

chromIntensityPlot(blData, scan.ids = 284, chrom.ids = 21)

# How would you detect trisomy 21 (i.e., describe briefly in words what the BAF/LRR plot would look like).

print("I would expect the BAF plot to have 4 distinct bands around 0, 0.333, 0.667, and 1.")
print("I would expect the band of the LRR plot to be upshifted form 0 to around 0.5 or higher.")

# Suppose an individual had a balanced reciprocal translocation involving chromosome 2 and 4. What would you expect that individual's BAF/LRR plots would look like for those chromosomes?

print("I would expect them to still appear normal. The balanced and recipricol nature of them makes the amount of genetic material and karyotypes unchanged, so I woulnd't expect to see any anomolies in the plots related to this.")

# Suppose chromosome 22 in sample 286 was the only chromosome flagged for a possible anomaly in this dataset. Looking at the plot, do you notice anything strange? Can you suggest what kind of anomaly might be showing up? (There isn't any single answer that's certainly right! The purpose of this question is to check that you understand what the BAF and LRR plots "mean").

print("It looks mostly fine. There may some signs for deletions and/or runs of homozygosity.")

# Finally, suppose your pipeline has given you a list of the exact chromosomal segments showing possibly aneuploidies/anomalies in each sample. What would you recommend doing with the genotypes in those segments? (You're not being asked to implement this here - just say what you recommend doing.)

print("Quite simply, I would recommend removing those genotypes from those chromosomal segments.")

```

## Easter Egg - Uncalled SNP Pokedex

```{r}
# references stackoverflow for this: https://stackoverflow.com/questions/73535956/how-to-insert-an-image-from-a-url-into-r-markdown-html-vignette
# and this:
# https://r-graph-gallery.com/363-add-images-and-links-in-kableExtra.html
get_pokemon_url <- function(number) {
  knitr::include_graphics(
    paste0(
      "https://raw.githubusercontent.com/PokeAPI/sprites/refs/heads/master/sprites/pokemon/",
      number,
      ".png"
    )
  )
}

print("Generating uncalled SNP Pokedex...")
library(kableExtra)
uncalled_snps <- snpAnnot %>% 
  getAnnotation() %>% 
  filter(missing.n1 == 1) %>%
  mutate(
    id = row_number()
  ) %>% 
  mutate(
    pokemon = get_pokemon_url(id)
  ) %>% 
  select(id, rsID, chromosome, pokemon)
pokemon <- uncalled_snps$pokemon
uncalled_snps %>% 
  mutate(pokemon = "") %>% 
  kbl(booktabs = T, align = "c") %>%
  kable_styling() %>%
  kable_paper(full_width = T) %>%
  column_spec(
    4,
    image = spec_image(pokemon, 280, 200) # dimensions of the images
  )

```
